MITM: Logging a Bluetooth Keyboard

### 2014-02-16 final_paper/paper.tex
added paper

\begin{abstract}
Despite the security improvements made to Bluetooth since its debut in the late 1990s, Bluetooth technology still has many unresolved vulnerabilities that allow for attackers to execute successful Man in the Middle (MITM) attacks. In particular, the addition of Secure Simple Pairing (SSP) in version 2.1 in 2007 introduced more security flaws, despite it's purpose to secure the pairing process between devices. This work focuses on launching a successful MITM attack against a Bluetooth-enabled keyboard and a victim PC and logging the data transmitted from the keyboard to the victim PC. Although there is much work done on MITM attacks and attacking PIN vulnerabilities, there has been little published research on successfully logging the keystrokes from a Bluetooth keyboard using a MITM attack. 
\end{abstract}

@@

\section{Introduction}
Bluetooth technology has proven to be vulnerable since it's adoption in to mainstream devices \cite{scarfone2008guide}, and while the security continues to improve, the threat for Man in the Middle (MITM) attacks against two paired Bluetooth devices remains high \cite{sandhya2012analysis}. In particular, MITM attacks take advantage of the Secure Simple Pairing (SSP) security protocol that was implemented in v2.1 of the Bluetooth stack during 2007, which was designed to enhance the security of Bluetooth devices and to simplify the pairing process for end-users. However, our work, and much of the previous work that focused on SSP MITM attacks, strives to show that there are still inherent security flaws in the implementation of the Bluetooth protocol.

@@

The rest of the paper is structured as follows. First, we discuss the background of Bluetooth. Then we move on to show the related work in the field we found. Next we present our threat model. We then move into discussing the attack and how each piece fits together. Then we discuss how we executed the attack and what we used to implement it. We then present our results and some ideas for how to mitigate any MITM attacks against Bluetooth keyboards. Lastly, we discuss future work and our conclusions.

@@

\subsection{Bluetooth}
Bluetooth is a short-range communications technology used in a wide variety of devices \cite{bluetooth}. The devices Bluetooth are implemented on varies widely from keyboards to fully functional computers \cite{bluetooth}. When two Bluetooth enabled devices are close by each other they can perform what is known as a pairing to connect themselves \cite{bluetooth}. Once the devices are connected they form a piconet and can communicate wirelessly \cite{bluetooth}. Any single Bluetooth device can connect with seven piconets at one time and inside any given piconet there can be up to seven Bluetooth devices connected \cite{bluetooth}.

\subsection{Bluetooth Security}
Bluetooth implements a variety of security measures to protect itself. Unfortunately, sometimes Bluetooth ends up defeating its own security measures. The following list are the major security features implemented currently.

\textbf{Frequency hopping}: The Bluetooth protocol cycles through a wide range of frequencies to avoid conflicts and collisions with other Bluetooth devices \cite{ma2007keystroke}. This cycling make it harder to capture packets from a remote host without having special hardware to scan the entire frequency spectrum. Special hardware like the Ubertooth dongle can map out the full spectrum and remotely sniff packets.

\textbf{Discoverable modes}: Devices that implement the Bluetooth standards can choose to only broadcast their availability for limited durations \cite{ma2007keystroke}. Having a hidden device can prevent against novice attackers from obtaining the Bluetooth address of the hidden device. This can be circumvented by performing a brute force attack on the address space and guessing the hidden devices Bluetooth address \cite{ma2007keystroke}.

\textbf{PIN authentication}: To pair two Bluetooth devices together there are several authentication options. One of the most secure options is when a PIN is used on both devices to confirm who the other person is \cite{ma2007keystroke}. Sadly, there exists a mode where no-PIN is needed that can be defaulted to thereby removing any security protections. Also, even if a PIN is used it is only numeric characters which makes it easy to crack \cite{ma2007keystroke}.

\textbf{Security Modes}: Bluetooth devices can be in a variety of different security modes with varying levels of security \cite{ma2007keystroke}. For backwards compatibility reasons the lowest level of security can still be used.
\begin{enumerate}
\item Mode 1: no security or encryption
\item Mode 2: no security until channel is established on L2CAP level (software)
\item Mode 3: has security before link setup on the LMP level (hardware)
\item Mode 4: introduced Secure Simple Pairing, pairing protocol to protect devices while a pairing is completed
\end{enumerate}

@@

\textbf{Numeric Comparison}: a 6 digit number is displayed on both devices and a user confirms that the numbers match (this is designed for devices that both have display and input capabilities)\\\\
\textbf{Passkey Entry}: designed when one device has input (keyboard) and the other mostly display; the device that has the display shows a 6 digit number, which the user types in on the input device.\\\\
\textbf{Just Works}: one device has no input or display; the user must accept the connection without being able to visualize that the numbers are the same\\\\
\textbf{Out of Band (OOB)}: designed to connect when two devices are touched together, (e.g. Near Field Communication or NFC)\\\\

@@

Among the vulnerabilities the National Institute of Standards and Technology (NIST) lists for Bluetooth versions before 4.0 is the weakness of one-way-only challenge and response authentication used in SSP \cite{scarfone2008guide}. 
Since Bluetooth devices often do not have display or input capabilities, Passkey Entry or Just Works modes for SSP are necessary to connect less-functional devices, such as keyboards, to more functional devices such as computers or tablets. When only one device is asking for authentication, it's possible to spoof the keyboard into pairing into Just Works mode and impersonate the keyboard to the victim computer. Despite NIST's statement that the one-way-only challenge affects versions before 4.0, if an attacker can force a keyboard into Just Works mode, it becomes vulnerable to MITM attacks.

\subsubsection{Pin Vulnerabilities}
NIST recommends that users use a 6-digit random pin when authenticating either in Numeric Comparison or Passkey Entry mode, however, this feature is not mandated. Some operating systems may allow for fewer than 6-digit pins, which are less secure, or let users enter custom pins, which will not be random. Assuming developers follow the recommended protocol, the probability of guessing the pin is one in a million according to NIST \cite{scarfone2008guide}.

Additionally, NIST claims that since the 6-digit passkey is not incorporated into the link key generation, it isn't useful for an eavesdropper to use. Nonetheless, Lindell showed in 2008 that the pin becomes useful to an attacker when attempting to pair with a device when the passkey is used more than once \cite{lindell2008attacks}. Knowing the pin can allow an attacker to connect with the device when the victim attempts to connect his/her devices again. It is possible to force a user to re-pair their devices either by a Denial of Service (DoS) attack \cite{cuthbertbluetooth} or by making the victims assume that the victim devices forgot the link key\cite{shaked2005cracking}. When the devices try to pair again, an attacker can intercept the process and authenticate itself to each device.

If the attacker does not know the pin, he/she can, in fact, crack it as Shaked and Wool showed in 2005 when they were able to crack a 4-digit pin in 0.27 seconds by observing Bluetooth traffic \cite{shaked2005cracking}. Similarly, in 2012, Barnickel, Wang, and Meyer were able to able to execute a MITM attack by first jamming the connection between the connected devices, calculate the passkey and then re-pairing with the devices \cite{barnickel2012implementing}. Both works and Lindell's work exploit the vulnerabilities of the Passkey Entry protocol, where the bit-by-bit evaluation of the passkey by devices can be used to leak the passkey a bit at a time, to determine the passkey quickly. As these works show, computing the pin is alarmingly simple and fairly useful when pins are reused; currently, this is still a feasible attack, but is less likely when pins are not fixed (such as with headsets or other fixed-pin devices).

\subsubsection{Keylogging}
Regardless of how infeasible it is to guess the password when it is not being reused, some successful approaches target Passkey Entry mode since attackers can force a user to enter the passkey for them. This is especially useful for devices that rely on Passkey Entry such as Bluetooth keyboards, which do not necessarily have display capabilities and use one-way authentication. Typically, Bluetooth enabled keyboards will connect to a device with display capabilities (e.g. a tablet or PC) and the device will display a passkey that the user must input through the connected keyboard. If an attacker can trick the user into entering the passkey for him/her, sidestepping the pin guessing process becomes quite easy.

@@

\section{Related Work}
MITM attacks are commonly researched for Bluetooth technologies and there are a variety of approaches to executing the attacks. As previously mentioned, the PIN prevents potential issues and is vulnerable to social engineering attacks. Phan and Mingard show that a malicious device can impersonate a legitimate device and communicate with another legitimate device by making the user enter whichever authentication pin it chooses \cite{phan2012analyzing}. Other approaches target reused pins by eavesdropping on the pairing process, jamming the signal, and then attempting to pair to the devices using the already-used pin \cite{barnickel2012implementing}. These approaches rely on the victim not thinking critically about the pin -- in some cases, the pin is hard-coded and so must be reused, or is set to default, or the user will use the same pin out of convenience. For this reason, these attacks rely on user-error, which does contribute largely to MITM exploits.

@@

\section{Threat Model}
In our threat model we are assuming the attack will be carried out in a public place where the attacker
can get within 30-100 feet of the victim. A couple examples include places like an airport, an office building, or a coffee shop. In any of these places it would be extremely easy for an attacker to get nearby the victim without appearing suspicious.

The victim of the attack is someone who is using a Bluetooth keyboard to type on a device. The devices could include a tablet, laptop, or PC. The victim would eventually in the course of their activities type some sensitive information. This could include payment information, passwords, or some other financial information.

The person performing the attack will most likely be using a laptop. They may or may not have specialty sniffing hardware. Their main goal will be attempting to log passwords or financial information like credit card numbers. To perform their attack they would most likely sit inconspicuously nearby the victim or hide out in a van outside the public place the attack is being carried out at.

\section{Attack Overview}
Based on prior research, the theoretical attack would take 4 main steps to executing the attack, which are as follow:
\begin{enumerate}
\item Find Bluetooth Address of the keyboard
\item Execute Denial of Service attack against victim
\item Pair with keyboard and victim
\item Forward keystrokes to victim
\end{enumerate}

@@

\subsection{Finding the Keyboard Bluetooth Address}
Obtaining a device's Bluetooth address is quite simple. If the device is in discoverable mode, then the address is easily found using a Bluetooth scanner or command line tool such as \texttt{hcitool scan} from the BlueZ Linux utilities. If the device is not in discoverable mode, but is connected to another Bluetooth-enabled device, then the address can be found using a Bluetooth sniffer and packet capturing software such as Kismet\cite{kismet}. Project Ubertooth \cite{ubertooth} provides both hardware and a plugin for Kismet that allows you to promiscuously sniff Bluetooth packets and then displays the Bluetooth address from whatever device is sending out the packets. Without pairing to the device, it is possible to sniff the packets from nearby devices, although the data is likely encrypted, and obtain the address \cite{hak5}.

In order to execute the attack, the attacker will need to know the Bluetooth address for both the keyboard and the victim PC. If the attacker does not have access to hardware like Ubertooth One, he or she can obtain the Bluetooth address after performing a DoS attack, at which point the user may make the devices discoverable in order to reconnect the devices and the attacker can scan for the devices.

\subsection{Denial of Service Attack}
The purpose of the Denial of Service attack (DoS) is to trick the user into initiating pairing. Although devices, once paired, do not technically need to be re-paired, a user who is frustrated with Bluetooth not connecting may try to pair the devices again to re-initiate the connection. Ultimately, this means the devices become discoverable to the attacker, who swiftly swoops in to pair with both devices.

There are several known approaches to doing a DoS attack against Bluetooth and DoS attacks present a realistic problem for Bluetooth users \cite{bandyopadhyay2010information}; an attacker can force some operating systems to disconnect with a keyboard by spamming the victim PC with link requests \cite{cuthbertbluetooth}. This attack worked for Windows Vista and Linux Ubuntu with BlueZ but did not work for Windows 7; as a result, if the attacker does not know what OS the user has installed, this approach may not be effective.

@@

\subsection{Getting in the Middle}
What makes this attack possible is the unreliability of Bluetooth connections. Users often experience problems connecting devices and understand that disconnections between their devices are common, so service interruptions are not out of the ordinary and are not suspicious. To fix connections, as in the case of a DoS attack, users will often try to re-pair their device or toggle the connect switch or button to try and connect the devices again. Once the user makes his or her devices discoverable, the attacker can quickly pair with both devices. The process must be executed quickly so that the victim keyboard does not have time to connect to the victim PC.

@@

The second step is to change the BlueZ configuration file to mimic the appearance of the user's keyboard. When the keyboard becomes discoverable \texttt{hcitool scan} can give you the name of the device, which the attacker simply change in the Bluetooth config file (found in \texttt{/etc/bluetooth/main.conf} for Ubunutu 12.04). The attacker must also change the device class so that the device will show up as a keyboard in the device manager of the victim PC. Typically this device id is \texttt{0x000540} for keyboards.

@@

Hidclient allows the user to forward from any input device the user specifies and that the computer recognizes as a valid input. Since the victim's Bluetooth keyboard is connected to the attacker already, hidclient recognizes the keyboard as an input source. This means the attacker's keyboard is free for typing and will not appear on the victim's screen. The keystrokes are forwarded on to the victim with minimal delay.

\subsection{Keylogging}
In order to get the data from the Bluetooth packets, we needed to dump the data in a way that it was retrievable either during the attack or after the fact. We considered several packet logger programs including Kismet \cite{kismet}, Wireshark \cite{wireshark}, and Mac OSX's Bluetooth Packet Logger. There were a few concerns with these tools: Kismet did not have support without the Ubertooth hardware, and so we did not choose to use this approach; Wireshark has Bluetooth support, but the data is difficult to extract, except through log files; and the OSX Packet Logger was only available on OSX, which meant we were unable to use Linux BlueZ utilities.

@@

To extract the data, we wrote a Python script based on findings of a Chinese blogger, alias Chopper \cite{chopper}. Chopper noted that data coming out of \texttt{hcidump} was not encrypted but encoded such that characters on the keyboard were given hexadecimal values that did not correspond to their ASCII values, as shown in Figure 1. However, decoding the output proved quite simple, and can be determined by simply pressing a key and looking at the corresponding \texttt{hcidump} output. We decoded some of them in our Python script to show a proof of concept, the values of which are shown in Table 1.

\section{Executing the Attack}
To execute this attack, we used completely off-the-shelf hardware and software. The only custom software and hardware would be the incorporation of the Ubertooth device and plugin for Kismet. 
\begin{table}
\begin{tabular}{|c|c|}
\hline
Key Value & hcidump Encoding\\\hline
A&0x04\\
B&0x05\\
C&0x06\\ 
D&0x07\\
E&0x08\\
F&0x09\\
G&0x0A\\
H&0x0B\\
I&0x0C\\
J&0x0D\\
K&0x0E\\
L&0x0F\\
M&0x10\\
N&0x11\\
O&0x12\\
P&0x13\\
Q&0x14\\
R&0x15\\
S&0x16\\
T&0x17\\
U&0x18\\
V&0x19\\
W&0x1A\\
X&0x1B\\
Y&0x1C\\
Z&0x1D\\
1&0x1E\\
2&0x1F\\
3&0x20\\
4&0x21\\
5&0x22\\
6&0x23\\
7&0x24\\
8&0x25\\
9&0x26\\\hline
\end{tabular}
\begin{center}
\caption{Keys to Encoded Value from \texttt{hcidump}}
\end{center}
\label{table1}
\end{table}

\subsection{Keylogging}
In order to get the data from the Bluetooth packets, we needed to dump the data in a way that it was retrievable either during the attack or after the fact. We considered several packet logger programs including Kismet \cite{kismet}, Wireshark \cite{wireshark}, and Mac OSX's Bluetooth Packet Logger. There were a few concerns with these tools: Kismet did not have support without the Ubertooth hardware, and so we did not choose to use this approach; Wireshark has Bluetooth support, but the data is difficult to extract, except through log files; and the OSX Packet Logger was only available on OSX, which meant we were unable to use Linux BlueZ utilities.

To extract the data, we wrote a Python script based on findings of a Chinese blogger, alias Chopper \cite{chopper}. Chopper noted that data coming out of \texttt{hcidump} was not encrypted but encoded such that characters on the keyboard were given hexadecimal values that did not correspond to their ASCII values, as shown in Figure 1. However, decoding the output proved quite simple, and can be determined by simply pressing a key and looking at the corresponding \texttt{hcidump} output. We decoded some of them in our Python script to show a proof of concept, the values of which are shown in Table 1.

@@

\textbf{Keyboard:} we used a basic Bluetooth keyboard that uses Bluetooth V3.0+EDR (Enhanced Data Rate (EDR), an optional speed-up of data transfer)\\
\textbf{Attack Device:} Lenovo Thinkpad z61t laptop running Ubuntu 12.04 with Bluetooth 2.0+EDR\\
\textbf{Victim Devices:} we attacked a mid-2012 Apple Macbook Air running Ubuntu 12.10 via VirtualBox with
Bluetooth 4.0 card and a desktop PC with a Bluetooth dongle, running Ubuntu 12.04 and Bluetooth 4.0.

To execute pairing, the Macbook, the Thinkpad Laptop and Tablet all use Passkey Entry authentication where a pin must be entered into the keyboard. Since the attack relies on SSP, which is largely the same across all versions of Bluetooth since it's implementation in v 2.1, the differences in Bluetooth versions across the devices should not matter. The largest variable in whether or not the attack is possible is how the OS handles the link keys and sending multiple link requests from a single address. 
\begin{figure}
\begin{verbatim}
ACL data: handle 11 flags 0x02 dlen 14
L2CAP(d): cid 0x0041 len 10 [psm 0]
A1 01 00 00 04 00 00 00 00 00
\end{verbatim}
\caption{Sample \texttt{hcidump} dump data (0x04) showing capital A}
\label{fig:mypic}
\end{figure}

\subsection{Setup}
Our attack setup used two laptops mentioned in the Hardware section and the Linux Command line utilities mentioned in the previous section. We used the Lenovo laptop as the attacker PC and the Macbook Air as the victim. Since the attack is OS ambivalent, the attack could theoretically be executed on a variety of Bluetooth enabled devices. However, given out limited resources, we were only able to attack the Macbook Air and another Desktop running Ubuntu 12.04 successfully.

@@

The attacker then runs the \texttt{hidclient} with the \texttt{-l} option, which lists the available inputs that can be forwarded to the victim PC.

@@

The Bluetooth keyboard appears as one of the options, and the attacker must simply replace the X in the following command to specify the input source as the keyboard.

@@

\section{Results}
We successfully created a proof of concept implementation that can forward keystrokes from a Bluetooth keyboard through the attacker to a victim machine. The proof of concept currently only works in the ideal conditions.

>>>>

\begin{abstract}
Despite the security improvements made to Bluetooth since its debut in the late 1990s, Bluetooth technology still has many unresolved vulnerabilities that allow for attackers to execute successful Man in the Middle (MITM) attacks. In particular, the addition of Secure Simple Pairing (SSP) in version 2.1 introduced more security flaws, despite it's purpose to secure the pairing process between devices. This work focuses on launching a successful MITM attack against a Bluetooth-enabled keyboard and a victim PC and logging the data transmitted from the keyboard to the victim PC. Although there is much work that has been done on MITM attacks and attacking PIN vulnerabilities, there has been no published research on successfully logging the keystrokes from a Bluetooth keyboard using a MITM attack. 
\end{abstract}

@@

\section{Introduction}
Bluetooth technology has proven to be vulnerable since it's adoption in to mainstream devices \cite{scarfone2008guide}, and while the security continues to improve, the threat for Man in the Middle (MITM) attacks against two paired Bluetooth devices remains high \cite{sandhya2012analysis}. In particular, MITM attacks take advantage of the Secure Simple Pairing (SSP) security protocol that was implemented in version 2.1 of the Bluetooth stack during 2007, which was designed to enhance the security of Bluetooth devices and to simplify the pairing process for end-users. However, our work, and much of the previous work that has focused on SSP MITM attacks, strives to show that there are still inherent security flaws in the implementation of the Bluetooth protocol.

@@

The rest of the paper is structured as follows. We discuss the background of Bluetooth and Bluetooth security in Section 2. In Section 3, we present the related work in the field regarding SSP vulnerabilities. In Section 4, we present our threat model. Section 5 covers the details of our attack model and implementation options. Section 6 covers the particular implementation and execution of our attack. In Section 7, we present our results and some ideas for how to mitigate any MITM attacks against Bluetooth keyboards in Section 8. Lastly, we discuss future work and our conclusions in Sections 10 and 11, respectively.

@@

\subsection{Bluetooth}
Bluetooth is a short-range communications technology that is used in a wide variety of devices \cite{bluetooth}. The types of devices that use Bluetooth technology vary from keyboards to fully functional computers \cite{bluetooth}. When two Bluetooth enabled devices are nearby one another, they can perform a pairing process to connect and enable data transfer \cite{bluetooth}. Once the devices are connected, they form a piconet and can communicate wirelessly \cite{bluetooth}. Any single Bluetooth device can connect with seven piconets at one time and inside any given piconet there can be up to seven Bluetooth devices connected \cite{bluetooth}.

\subsection{Bluetooth Security}
Bluetooth implements a variety of security measures to protect itself. The following list are the major security features implemented in the latest version of Bluetooth:

\textbf{Frequency hopping}: The Bluetooth protocol cycles through a wide range of frequencies to avoid conflicts and collisions with other Bluetooth devices \cite{ma2007keystroke}. While not a security feature, frequency cycling make it harder to capture packets from a remote host without having special hardware to scan the entire frequency spectrum. However, special hardware like the Ubertooth dongle can map out the full spectrum and remotely sniff packets.

\textbf{Discoverable modes}: Bluetooth devices can choose to only broadcast their availability and Bluetooth address for limited durations \cite{ma2007keystroke}. Hiding the device limits the risk of other devices discovering it and trying to connect to it, although devices can remain always discoverable. Additionally, attackers can still obtain the Bluetooth address by performing a brute force attack on the address space \cite{ma2007keystroke}.

\textbf{PIN authentication}: To pair two Bluetooth devices together there are several authentication options. One of the more secure options is when a PIN is used on both devices to authenticate the devices that the user is pairing \cite{ma2007keystroke}. Not all modes require two-way authentication; there exists a mode where no-PIN is needed or a default or static pin is used, thereby removing any security protections. Also, even if a PIN is used, it is only numeric characters which makes it easy to crack \cite{ma2007keystroke,shaked2005cracking,lindell2008attacks}.

\textbf{Security Modes}: To deal with back-compatibility as well as less-functional devices such as Bluetooth receivers, Bluetooth implements 4 security modes with varying levels of security \cite{ma2007keystroke}. For backwards compatibility, the lowest level of security can still be used.
\begin{enumerate}
\item Mode 1: no security or encryption
\item Mode 2: no security until channel is established on L2CAP level (software)
\item Mode 3: has security before link setup on the LMP level (hardware)
\item Mode 4: introduced Secure Simple Pairing, pairing protocol to protect devices while a pairing is completed
\end{enumerate}

Most PCs use Mode 4, which enables the SSP process and provides encryption; however, peripheral devices like keyboards and mice, which support Mode 4, can still connect to PCs and use Mode 1 for transferring data. This exposes the data to anyone sniffing the traffic.

@@

\textbf{Numeric Comparison}: a 6 digit number is displayed on both devices and a user confirms that the numbers match (this is designed for devices that both have display and input capabilities)\\\\
\textbf{Passkey Entry}: designed for when one device has input (keyboard) and the other has display capabilities; the device that has the display shows a pin number, which the user types in on the input device.\\\\
\textbf{Just Works}: designed for when one device has no input or display capabilities; the user must accept the connection without being able to visually confirm that the pin numbers are the same\\\\
\textbf{Out of Band (OOB)}: designed to connect when two devices are touched together, (e.g. Near Field Communication or NFC)\\

\textit{Passkey Entry} mode is the primary mode of pairing keyboards and computers; however, keyboards can connect in \textit{Just Works} mode, which bypasses the security features of SSP \cite{4401672}.

@@

Among the vulnerabilities the National Institute of Standards and Technology (NIST) lists for Bluetooth versions before 4.0 is the weakness of one-way-only challenge and response authentication used in SSP \cite{scarfone2008guide}. Since Bluetooth devices often do not have display or input capabilities, \textit{Passkey Entry} or \textit{Just Works} modes for SSP are necessary to connect less-functional devices, such as keyboards, to more functional devices such as computers or tablets. When only one device is asking for authentication, it's possible to spoof the keyboard into pairing into \textit{Just Works} mode and impersonate the keyboard to the victim computer. Despite NIST's statement that the one-way-only challenge affects versions before 4.0, if an attacker can force a keyboard into \textit{Just Works} mode or use social-engineering or brute-force pin attacks in \textit{Passkey Entry} mode, the device becomes vulnerable to MITM attacks.

\subsubsection{Pin Vulnerabilities}
NIST recommends that users use a 6-digit random pin when authenticating either in Numeric Comparison or Passkey Entry mode, however, this feature is not mandated. Some operating systems may allow for fewer than 6-digit pins, which are less secure, or may let users enter custom pins, which will not be random and may be less secure. Assuming developers follow the recommended protocol, the probability of guessing the pin is one in a million according to NIST \cite{scarfone2008guide}.

NIST also claims that since the 6-digit passkey is not incorporated into the link key generation, it isn't useful for an eavesdropper to use. Nonetheless, Lindell showed in 2008 that the pin becomes useful to an attacker when attempting to pair with a device when the passkey is used more than once \cite{lindell2008attacks}. Knowing the pin can allow an attacker to connect with the device when the victim attempts to connect his/her devices again. It is possible to force a user to re-pair their devices either by a Denial of Service (DoS) attack \cite{cuthbertbluetooth} or by making the victims assume that the victim devices forgot the link key\cite{shaked2005cracking}. When the devices try to pair again, an attacker can intercept the process and authenticate itself to each device.

If the attacker does not know the pin, he/she can, in fact, crack it as Shaked and Wool showed in 2005 when they were able to crack a 4-digit pin in 0.27 seconds by observing Bluetooth traffic \cite{shaked2005cracking}. Similarly, in 2012, Barnickel, Wang, and Meyer were able to able to execute a MITM attack by first jamming the connection between the connected devices, calculate the passkey and then re-pairing with the devices \cite{barnickel2012implementing}. Both of these works and Lindell's work exploit the vulnerabilities of the \textit{Passkey Entry} protocol, where the bit-by-bit evaluation of the passkey by devices can be used to leak the passkey a bit at a time, to determine the passkey quickly. As these works show, computing the pin is alarmingly simple and fairly useful when pins are reused; currently, this is still a feasible attack, but is less likely when pins are not fixed (such as with headsets or other fixed-pin devices).

\subsubsection{Keylogging}
Regardless of how infeasible it is to guess the password when it is not being reused, some successful approaches target \textit{Passkey Entry} mode since attackers can force a user to enter the passkey for them. This is especially useful for devices that rely on one-way authentication such as Bluetooth keyboards, which do not necessarily have display capabilities. Typically, Bluetooth enabled keyboards will connect to a device with display capabilities (e.g. a tablet or PC) and the device will display a passkey that the user must input through the connected keyboard. If an attacker can trick the user into entering the passkey for him/her, they sidestep the pin guessing process and the attack becomes quite easy.

@@

\section{Related Work}
MITM attacks are commonly researched for Bluetooth technologies and there are a variety of approaches to executing the attacks. As previously mentioned, the PIN prevents potential issues and is vulnerable to social engineering attacks. Phan and Mingard show that a malicious device can impersonate a legitimate device and communicate with another legitimate device by making the user enter whichever authentication pin it chooses \cite{phan2012analyzing}. Other approaches target reused pins by eavesdropping on the pairing process, jamming the signal, and then attempting to pair to the devices using the already-used pin \cite{barnickel2012implementing,lindell2008attacks}. These approaches rely on the victim not thinking critically about the pin -- in some cases, the pin is hard-coded and so must be reused, or is set to default, or the user will use the same pin out of convenience. For this reason, these attacks rely on user-error, which is a largely used resource for MITM exploits.

@@

\section{Threat Model}
In our threat model, we are assuming the attack will be carried out in a public place where the attacker
can get within 30-100 feet of the victim. Examples include places like an airport, an office building or a coffee shop. In any of these places, it would be extremely easy for an attacker to get nearby the victim without appearing suspicious.

The victim of the attack is someone who is using a Bluetooth keyboard to type on a device. The devices could include a tablet, laptop, or desktop PC. The victim would eventually type some sensitive information. This could include payment information, passwords, or some other financial information.

The person performing the attack will most likely be using a laptop. The attacker may or may not have speciality sniffing hardware. His or her main goal will be attempting to log passwords or financial information like credit card numbers. To perform the attack, the attacker would most likely sit inconspicuously nearby the victim or hide out in a van outside the public place where the attack is being carried out.

\section{Attack Overview}
Based on prior research, the theoretical attack would take 4 main steps to execute, which are as follow:
\begin{enumerate}
\item Find Bluetooth Address of the keyboard
\item Execute Denial of Service attack against victim
\item Pair with keyboard and victim
\item Forward keystrokes to victim
\end{enumerate}

@@

\subsection{Finding the Keyboard Bluetooth Address}
Obtaining a device's Bluetooth address is quite simple. If the device is in discoverable mode, then the address is easily found using a Bluetooth scanner or command line tool such as \texttt{hcitool scan} from the BlueZ Linux utilities. If the device is not in discoverable mode but is connected to another Bluetooth-enabled device, then the address can be found using a Bluetooth sniffer and packet capturing software such as Kismet\cite{kismet}. Project Ubertooth \cite{ubertooth} provides both hardware and a plugin for Kismet that allows you to promiscuously sniff Bluetooth packets, and displays the Bluetooth address from whatever device is sending out the packets. Without pairing to the device, it is possible to sniff the packets from nearby devices, although the data is likely encrypted \cite{hak5}.

In order to execute the attack, the attacker will need to know the Bluetooth address for both the keyboard and the victim PC. If the attacker does not have access to hardware like Ubertooth One, he or she can obtain the Bluetooth address after performing a DoS attack, at which point the user may make the devices discoverable in order to reconnect the devices and the attacker can scan for the device addresses.

\subsection{Denial of Service Attack}
Denial of Service (DoS) attacks are powerful attacks that can disrupt device function or cause harm to the device \cite{premnath2008battery}. The purpose of the DoS attack with our approach is to trick the user into initiating pairing. Although devices, once paired, do not technically need to be re-paired, a user who is frustrated with Bluetooth not connecting may try to pair the devices again to re-initiate the connection. Ultimately, this means the devices become discoverable to the attacker, who swiftly swoops in to pair with both devices.

There are several known approaches to executing a DoS attack against Bluetooth, present a realistic and serious problem for Bluetooth users \cite{bandyopadhyay2010information}; an attacker can force some operating systems to disconnect with a keyboard by spamming the victim PC with link requests \cite{cuthbertbluetooth}. This attack worked for Windows Vista and Linux Ubuntu with BlueZ but did not work for Windows 7; as a result, if the attacker does not know what OS the user has installed, this approach may not be effective.

@@

\subsection{Getting in the Middle}
What makes this attack possible is the unreliability of Bluetooth connections. Users often experience problems connecting devices and understand that disconnections between their devices are common, so service interruptions are not out of the ordinary and are not suspicious. To fix connections, as in the case of a DoS attack, users will often try to re-pair their device or toggle the physical connect switch or button to try and connect the devices again. Once the user makes his or her devices discoverable, the attacker can quickly pair with both devices. The process must be executed quickly so that the victim keyboard does not have time to connect to the victim PC.

@@

The second step is to change the BlueZ configuration file to mimic the appearance of the user's keyboard. When the keyboard becomes discoverable, \texttt{hcitool scan} can give you the name of the device, which the attacker simply can change in the Bluetooth config file (found in \texttt{/etc/bluetooth/main.conf} for Ubunutu 12.04). The attacker must also change the device class so that the device will show up as a keyboard in the device manager of the victim PC. Typically this device id is \texttt{0x000540} for keyboards.

@@

Hidclient allows the user to forward from any input device the user specifies and that the computer recognizes as a valid input. Since the victim's Bluetooth keyboard is connected to the attacker already, hidclient recognizes the keyboard as an input source. This means the attacker's keyboard is free for typing and will not appear on the victim's screen. The keystrokes are forwarded on to the victim with minimal delay. 
\begin{table}
\begin{tabular}{|c|c|}
\hline
Key Value & hcidump Encoding\\\hline
A&0x04\\
B&0x05\\
C&0x06\\ 
D&0x07\\
E&0x08\\
F&0x09\\
G&0x0A\\
H&0x0B\\
I&0x0C\\
J&0x0D\\
K&0x0E\\
L&0x0F\\
M&0x10\\
N&0x11\\
O&0x12\\
P&0x13\\
Q&0x14\\
R&0x15\\
S&0x16\\
T&0x17\\
U&0x18\\
V&0x19\\
W&0x1A\\
X&0x1B\\
Y&0x1C\\
Z&0x1D\\
1&0x1E\\
2&0x1F\\
3&0x20\\
4&0x21\\
5&0x22\\
6&0x23\\
7&0x24\\
8&0x25\\
9&0x26\\\hline
\end{tabular}
\begin{center}
\caption{Keys to Encoded Value from \texttt{hcidump}}
\end{center}
\label{table1}
\end{table}
\subsection{Keylogging}
In order to get the data from the Bluetooth packets, we needed to dump the data in a way that it was retrievable either during the attack or after the fact. We considered several packet logger programs including Kismet \cite{kismet}, Wireshark \cite{wireshark}, and Mac OSX's Bluetooth Packet Logger. There were a few concerns with these tools: Kismet did not have support without the Ubertooth hardware, and so we did not choose to use this approach; Wireshark has Bluetooth support, but the data is difficult to extract, except through log files; and the OSX Packet Logger was only available on OSX, which meant we were unable to use Linux BlueZ utilities.

@@

To extract the data, we wrote a Python script based on findings of a Chinese blogger, aliased Chopper \cite{chopper}. Chopper noted that data coming out of \texttt{hcidump} was not encrypted but encoded such that characters on the keyboard were given hexadecimal values that did not correspond to their ASCII values, as shown in Figure 1. However, decoding the output proved quite simple, and can be determined by simply pressing a key and looking at the corresponding \texttt{hcidump} output. We decoded some of them in our Python script to show a proof of concept, the values of which are shown in Table 1.

@@

\textbf{Keyboard:} we used a basic Bluetooth keyboard that uses Bluetooth V3.0+EDR (Enhanced Data Rate (EDR), an optional speed-up of data transfer)\\
\textbf{Attack Device:} we used a Lenovo Thinkpad z61t laptop running Ubuntu 12.04 with Bluetooth 2.0+EDR\\
\textbf{Victim Devices:} we attacked a mid-2012 Apple Macbook Air running Ubuntu 12.10 via VirtualBox with
Bluetooth 4.0 card and a desktop PC with a Bluetooth dongle, running Ubuntu 12.04 and Bluetooth 4.0.

To execute pairing, the Macbook, the Thinkpad Laptop and Tablet all use \textit{Passkey Entry} authentication, where a pin must be entered into the keyboard. Since the attack relies on SSP, which is largely the same across all versions of Bluetooth since it's implementation in version 2.1, the differences in Bluetooth versions across the devices should not matter. The largest variable in whether or not the attack is possible is how the OS handles the DoS attack.
\begin{figure}
\begin{verbatim}
ACL data: handle 11 flags 0x02 dlen 14
L2CAP(d): cid 0x0041 len 10 [psm 0]
A1 01 00 00 04 00 00 00 00 00
\end{verbatim}
\caption{Sample \texttt{hcidump} dump data (0x04) showing capital A}
\label{fig:mypic}
\end{figure}

\subsection{Setup}
Our attack setup used two laptops mentioned in the Hardware section and the Linux Command line utilities mentioned in the previous section. We used the Lenovo laptop as the attacker PC and the Macbook Air as the victim. Since the attack is theoretically OS ambivalent, the attack could theoretically be executed on a variety of Bluetooth enabled devices. However, given out limited resources, we were only able to attack the Macbook Air and another Desktop running Ubuntu 12.04 successfully.

@@

The attacker then runs the \texttt{hidclient} with the \texttt{-l} option, which lists the available inputs that can be forwarded to the victim PC:

@@

The Bluetooth keyboard appears as one of the options from the \texttt{-l} output, and the attacker must simply replace the X in the following command to specify the input source as the keyboard.

@@

\section{Results}
We successfully created a proof of concept implementation that can forward keystrokes from a Bluetooth keyboard through the attacker to a victim machine. The proof of concept currently only works in the ideal conditions. However, this approach is superior to simple Bluetooth packet sniffing, because it ensures that the data being read from the keyboard is unencrypted, which is not usually the case for most paired keyboards.
